---
title: "Homework 4 PLS 298"
author: "Elena Suglia"
date: "12/11/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
source("plots.R")
```

setwd("~/Box Sync/Graduate School/PhD Research/Ch. 2) Lassen Snowmelt & Phenology/Data analysis/lassen-phenology")
source("~/Box Sync/Graduate School/PhD Research/Ch. 2) Lassen Snowmelt & Phenology/Data analysis/lassen-phenology/plots.R")

## The relationship between environment and phenology in the native California wildflower *Streptanthus tortuosus* at Lassen Volcanic National Park

This work represents preliminary analyses for a chapter of my dissertation proposal. In the following analyses, I'm investigating variation in phenology of *Streptanthus tortuosus* plants in relation to environmental variables at Lassen Volcanic National Park along an elevation gradient. I collected data every 7-10 days throughout the growing season from 2017-2019 at four populations with varying numbers of plots within those populations and varying numbers of marked individuals within those plots (ranging from 20-100, but fewer in some years). Plants at these populations are biennials or iteroparous perennials that germinate in the spring, grow into vegetative rosettes their first year and overwinter until spring the next year, when they melt out and flower. This data includes number of buds, flowers, and fruits on each plant every week, as well as some size and morphological metrics taken less frequently like height, stem diameter, number of branches, canopy width, and leaf size. I also measured temperature every 3 hours in 2018 and 2019 using small temperature sensors (iButtons) installed in the field (more were installed in 2019 than in 2018). The temperature data also provides me with metrics related to snowpack, an important abiotic factor in alpine environments. These data include snowmelt timing and length of snow-free period.

One thing to note here is that snowmelt timing does not vary linearly with elevation, which offers an opportunity to decouple snowmelt, elevation, and phenology in ways that are usually not feasible:

```{r, echo = FALSE, warning = FALSE, comment = FALSE}
elsnowmelt
```

* Multiple points for some populations in some years indicate several plots.

Let's also visualize the temperature data:

```{r, echo = FALSE, warning = FALSE, comment = FALSE}
tempelev

monthvstemp
```

The specific questions I would like to answer in this chapter are:

1. What is the relationship between environment and phenology across years and populations?

2. Do fitness and selection vary in relation to phenology and environment?

3. How does the environment relate to synchronization of flowering across populations?

I realize the term "environment" is rather vague. This is the order of priority of environmental variables I'm interested in investigating, and I will complete as many of them as I can before my dissertation proposal is due:

* snowmelt timing
* temperature
* growing degree days
* photoperiod (daylength) accumulation
* photothermal accumulation

# 1. Relationship between environment and phenology

Let's visualize how populations are timing their flowering in relation to snowmelt timing:

```{r, warning = FALSE, comment = FALSE}
require(grid)
grid.arrange(d5plot18, d5plot19, nrow = 1, top=textGrob("Date vs percent population flowering"))

```

* Populations are ordered from highest elevation at the top to lowest at the bottom
* The x-axis is day of the year (Julian day)
* The y-axis is percent of the reproductive individuals in the population that are flowering
* The dashed gray vertical lined is snowmelt timing
* The green vertical line is "peak flowering," which is the day on which that population has the greatest total number of flowers
* The red vertical line is the last day that the population had any flowers (included to provide  a visualization of flowering duration between the green and red lines)
* I do not include first flowering in this figure because I missed it for several populations (it's too challening to reach certain populations before the snow melts out)
* Ignore LV3 in 2019 (bottom right population); it experienced a landslide halfway through the 2019 growing season that wiped out all my marked individuals

Here, we see that there is variation in snowmelt timing and flowering time across populations and years (note the different x axes in 2018 and 2019; the snowmelt timing and therefore the growing season were much delayed in 2019 compared to 2018). We also see what appears to be a relationship between snowmelt timing and days from snowmelt to peak flowering: the later the snow melts, the faster the population flowers.

### Relationship between flowering time and snowmelt date

Let's look at the relationship between flowering date as response variables and snowmelt date as the explanatory variable. Here, I include population and year as random variables, and plot as a random variable nested within population.

First, we can visualize the relationships:

```{r, warning = FALSE, comment = FALSE}
require(gridExtra)
grid.arrange(plotflrsnow18, plotflrsnow19, nrow=2)
```

There is a lot of individual variation, but later snowmelt appears to delay all metrics of flowering phenology.

Now let's do some statistical analyses. In the following, the day of year of snowmelt (Julian day) is denoted as "jmeltdate."


```{r, warning=FALSE, message = FALSE}
# peakflr ~ jmeltdate model selection

# peakflr ~ jmeltdate fixed effects model & anova

peakflrlm = lm(peakflr ~ jmeltdate, data = i)
summary(peakflrlm)
anova(peakflrlm)
# later snowmelt significantly delays peak flowering

# mixed models for peakflr ~ jmeltdate
m6 = lmer(peakflr ~ jmeltdate + (1|site), data = i, REML = FALSE)
m7 = lmer(peakflr ~ jmeltdate + (1 + jmeltdate|site), data = i, REML = FALSE)
m8 = lmer(peakflr ~ jmeltdate + (1|year), data = i, REML = FALSE)
m9 = lmer(peakflr ~ jmeltdate + (1|site) + (1|year), data = i, REML = FALSE)
m10 = lmer(peakflr ~ jmeltdate*year + (1|site/plot), data = i, REML = FALSE) # best fit
m11 = lmer(peakflr ~ jmeltdate*site + (1|year), data = i, REML = FALSE)
m12 = lmer(peakflr ~ jmeltdate*site + (1|year) + (1|site), data = i, REML = FALSE)
m13 = lmer(peakflr ~ jmeltdate*site + (1 + site|year), data = i, REML = FALSE)

AIC(peakflrlm, m6, m7, m8, m9, m10, m11, m12, m13)
AICc(peakflrlm, m6, m7, m8, m9, m10, m11, m12, m13)
BIC(peakflrlm, m6, m7, m8, m9, m10, m11, m12, m13)

# best fit model is model 10 and is consistent across model comparison methods


#r.squaredGLMM(m6)
#r.squaredGLMM(m7) 
#r.squaredGLMM(m8) 
#r.squaredGLMM(m9) 
r.squaredGLMM(m10) 
#r.squaredGLMM(m11) 
#r.squaredGLMM(m12) 
#r.squaredGLMM(m13) 

# including random effects improves all models; displayed results of best fit model only

# firstflr ~ jmeltdate model selection

# firstflr ~ jmeltdate fixed model & anova
firstflrlm = lm(firstflr ~ jmeltdate, data = i)
summary(firstflrlm)
anova(firstflrlm)
# later snowmelt significantly delays first flowering 

m14 = lmer(firstflr ~ jmeltdate + (1|site), data = i, REML = FALSE)
m15 = lmer(firstflr ~ jmeltdate + (1 + jmeltdate|site), data = i, REML = FALSE)
m16 = lmer(firstflr ~ jmeltdate + (1|year), data = i, REML = FALSE)
m17 = lmer(firstflr ~ jmeltdate + (1|site) + (1|year), data = i, REML = FALSE)
m18 = lmer(firstflr ~ jmeltdate*year + (1|site), data = i, REML = FALSE)
m19 = lmer(firstflr ~ jmeltdate*site + (1|year), data = i, REML = FALSE) # best fit AIC
m20 = lmer(firstflr ~ jmeltdate*site + (1|year) + (1|site), data = i, REML = FALSE)
m21 = lmer(firstflr ~ jmeltdate*site + (1 + site|year), data = i, REML = FALSE)
# some warning messages

AIC(firstflrlm, m14, m15, m16, m17, m18, m19, m20, m21)
# AICc(firstflrlm, m14, m15, m16, m17, m18, m19, m20, m21)
# BIC(firstflrlm, m14, m15, m16, m17, m18, m19, m20, m21) # BIC gives m18 as best fit model

# Only displaying the metrics for best fit model

#r.squaredGLMM(m14)
#r.squaredGLMM(m15) 
#r.squaredGLMM(m16) 
#r.squaredGLMM(m17) 
#r.squaredGLMM(m18) 
r.squaredGLMM(m19) 
#r.squaredGLMM(m20) 
#r.squaredGLMM(m21) 

# including random effects does not necessarily appear to improve model performance
# why are the r.squared values identical for this model? What do the messages mean?

# lastflr ~ jmeltdate model selection

# lastflr ~ jmeltdate fixed effects model
lastflrlm = lm(lastflr ~ jmeltdate, data = i)
summary(lastflrlm)
anova(lastflrlm)

# mixed models for meltdate ~ lastflr
m22 = lmer(lastflr ~ jmeltdate + (1|site), data = i, REML = FALSE)
m23 = lmer(lastflr ~ jmeltdate + (1 + jmeltdate|site), data = i, REML = FALSE)
m24 = lmer(lastflr ~ jmeltdate + (1|year), data = i, REML = FALSE)
m25 = lmer(lastflr ~ jmeltdate + (1|site) + (1|year), data = i, REML = FALSE)
m26 = lmer(lastflr ~ jmeltdate*year + (1|site), data = i, REML = FALSE) # best fit BIC
m27 = lmer(lastflr ~ jmeltdate*site + (1|year), data = i, REML = FALSE) # best fit AIC
m28 = lmer(lastflr ~ jmeltdate*site + (1|year) + (1|site), data = i, REML = FALSE)
m29 = lmer(lastflr ~ jmeltdate*site + (1 + site|year), data = i, REML = FALSE)

AIC(lastflrlm, m22, m23, m24, m25, m26, m27, m28, m29)
#AICc(lastflrlm, m22, m23, m24, m25, m26, m27, m28, m29)
#BIC(lastflrlm, m22, m23, m24, m25, m26, m27, m28, m29)
# BIC says model 26 is best

#r.squaredGLMM(m22)
#r.squaredGLMM(m23) 
#r.squaredGLMM(m24) 
#r.squaredGLMM(m25) 
#r.squaredGLMM(m26) 
r.squaredGLMM(m27) 
#r.squaredGLMM(m28) 
#r.squaredGLMM(m29) 

```

What about the relationship between flowering date and elevation?
Using peak flowering as the representative flowering phenology metric from now on.

```{r, warning = FALSE}
# peakflr ~ elevation
elevpeakflrlm = lm(peakflr ~ elevation, data = i)
summary(elevpeakflrlm)
anova(elevpeakflrlm)
# significant, but less so than snowmelt date
```

# flowering duration ~ jmeltdate ----
meltdatedurationlm = lm(flrduration ~ jmeltdate, data = i)
summary(meltdatedurationlm)
anova(meltdatedurationlm)
# this is a pretty interesting result. Flowering duration is more influenced when snowmelt is later (or is it something else about that year?)
# days from snowmelt to first flr ~ jmeltdate ----
meltdatesnowtoflrlm = lm(snowtoflr ~ jmeltdate, data = i)
summary(meltdatesnowtoflrlm)
anova(meltdatesnowtoflrlm)
AIC(meltdatesnowtoflrlm)
# flower much faster when snow melts later. This could either be developmental (growth is faster in warmer temps) or due to anticipatory/photoperiod cuing to synchronize flowering with other plants.

# ---------------------------------------------------------------------------
# 2.2 Relationships between environment, phenology, and fitness ----
# fitness ~ snowmelt date ----

# fixed effects model & anova
fitnesssnowlm = lm(fitness ~ jmeltdate, data = i)
summary(fitnesssnowlm)
anova(fitnesssnowlm)
# site is important. not sure how to interpret this result

# mixed model
m1 = lmer(fitness ~ jmeltdate + (1 + jmeltdate|year), data = i, REML = FALSE)
m2 = lmer(fitness ~ jmeltdate + (1|site), data = i, REML = FALSE)
m2.5 = lmer(fitness ~ jmeltdate + (1|jmeltdate), data = i, REML = FALSE)
m3 = lmer(fitness ~ jmeltdate + (1|year), data = i, REML = FALSE)
m4 = lmer(fitness ~ jmeltdate + (1 + jmeltdate|year) + (1|site), data = i, REML = FALSE)
m5 = lmer(fitness ~ jmeltdate*year +  (1|site), data = i, REML=FALSE)

display(m1)
display(m2)
AIC(fitnesssnowlm, m1, m2, m2.5, m3, m4, m5)
AICc(fitnesssnowlm, m1, m2, m2.5, m3, m4, m5)
BIC(fitnesssnowlm, m1, m2, m2.5, m3, m4, m5)

# Do we include the random effects or not?
# The marginal R2 value represents the variance explained by the fixed effects
# The conditional R2 value is interpreted as the variance explained by the entire model, including both fixed and random effects

r.squaredGLMM(m1) # R2m        R2c
# 0.0560748  0.1495234
r.squaredGLMM(m2) 
r.squaredGLMM(m2.5)
r.squaredGLMM(m3) 
r.squaredGLMM(m4) 
r.squaredGLMM(m5) 
# all models perform better when including random effects

# fitness ~ peakflr ----
fitnesspeakflrlm = lm(fitness ~ peakflr, data = i)
summary(flrlm)
anova(flrlm)

# fitness ~ firstflr ----
fitnessfirstflrlm = lm(fitness ~ firstflr, data = i)
summary(flrlm)
anova(flrlm)

# fitness ~ days from snowmelt to first flr ----
snowfirstflrlm = lm(fitness ~ snowtoflr*year, data = i)
summary(snowfirstflrlm)
anova(snowfirstflrlm)
# relationship between fitness and days from snowmelt to first flowering are totally dependent on year!


# fitness ~ branchiness ----
branchfitnesslm = lm(fitness ~ branchiness, data = i)
summary(branchfitnesslm)
anova(branchfitnesslm)
# Those with more (max) total branches had significantly higher fecundity.